Type theory in type theory with single substitutions

### 2024-09-18 types_2024_postproc/p.tex
stuff

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$ as follows.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers such as $\lam$. Notably,
we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
important to truncate the $\alpha$-normal predicates to be propositions to
ensure that it preserves the $\beta$/$\eta$ conversion rules.

@@

Instead of doing induction on $\alpha$-normal forms for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\sigma, \delta : \Sub^*\,\Delta\,\Gamma$, if $A[\sigma] =
  A[\delta]$ and $x[\sigma] = x[\delta]$ for any $A:\Ty\,\Gamma$ and variable
  $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\sigma] = \Omega[\delta]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\sigma^{+^\Omega}] = A[\delta^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\sigma^{+^\Omega}] = a[\delta^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\sigma^{+^\Omega}] =
  x[\delta^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}

>>>>

A type or term is $\alpha$-normal if it does not contain instantiation
operations, except at the leaves of the syntax tree as variables. However,
$\alpha$-normal terms can still contain $\beta$/$\eta$ redexes, so that we do
not need to do full normalization. We define variables and $\alpha$-normal forms
as inductive predicates in $\Prop$ as follows.
\begin{itemize}
  \item $\q:\Tm\,(\Gamma\ext A)\ (A[\p])$ is a variable.
  \item $x[\p]:\Tm\,(\Gamma\ext A)\ (B[\p])$ is a variable if $x:\Tm\,\Gamma\,B$
  is a variable.
  \item $x:\Tm\,\Gamma\,A$ is $\alpha$-normal if $x$ is a variable.
  \item $\Pi\,A\,B:\Ty\,\Gamma\,i$ is $\alpha$-normal if $A$ and $B$ are
  $\alpha$-normal types.
  \item $\app\,f\,a:\Tm\,\Gamma\,(B[\langle a \rangle])$ is $\alpha$-normal if
  $A$ and $B$ are $\alpha$-normal types, $f:\Tm\,\Gamma\,(\Pi\,A\,B)$ and
  $a:\Tm\,\Gamma\,A$ are $\alpha$-normal terms.
\end{itemize}
The predicate is defined similarly for the other type and term formers such as $\lam$. Notably,
we do not state that instantiated types/terms are $\alpha$-normal (except variables). It is
important to truncate the $\alpha$-normal predicates to be propositions to
ensure that they preserve the $\beta$/$\eta$ conversion rules.

@@

Instead of doing induction on $\alpha$-normal forms for each of the equations in
(\ref{eq:lifted}), we define a general lemma which can lift any equation between
instantiations over a telescope. For this we define $\Sub^*$ to be $\Sub$ with
freely added identity and composition operations, we do not require it to
satisfy the laws of a category as we will not compare $\Sub^*$s directly for
equality, only between types and terms instantiated with $\Sub^*$s. All
instantiation operations, lifting operations, and substitution rules are
redefined for $\Sub^*$.
\begin{lemma}
  \label{thm:lifting}
  Given $\gamma_0, \gamma_1 : \Sub^*\,\Delta\,\Gamma$, if $A[\gamma_0] =
  A[\gamma_1]$ and $x[\gamma_0] = x[\gamma_1]$ for any $A:\Ty\,\Gamma$ and
  variable $x:\Tm\,\Gamma\,A$, then
  \begin{itemize}
    \item $\Omega[\gamma_0] = \Omega[\gamma_1]$ for $\Omega:\Tel\,\Gamma$
    \item $A[\gamma_0^{+^\Omega}] = A[\gamma_1^{+^\Omega}]$ for $A:\Ty\,(\Gamma+\Omega)$
    \item $a[\gamma_0^{+^\Omega}] = a[\gamma_1^{+^\Omega}]$ for $a:\Tm\,(\Gamma+\Omega)\,A$
  \end{itemize}
  Note that the later equations depend the the earlier ones.
\end{lemma}
\begin{proof}
  Assuming the first equation, we prove that $x[\gamma_0^{+^\Omega}] =
  x[\gamma_1^{+^\Omega}]$ for any variable $x$ by induction on the telescope and
  the variable. Then the two latter equations can be proven by mutual induction
  on $\alpha$-normalized types and terms, still assuming the first equation.
  Finally we prove the first equation by induction on the telescope, using the
  previously proven equation for types, discharging its assumption.
\end{proof}
